---
title: "Analyze microarray data from Marks et al 2008"
output: html_notebook
editor_options: 
  chunk_output_type: inline
---

The script below are based on several tutorials, including
- the [manual](https://www.bioconductor.org/packages/devel/bioc/vignettes/limma/inst/doc/usersguide.pdf) for the LIMMA package, Section 17.4
- [Tutorial](https://www.bioconductor.org/packages/devel/workflows/vignettes/arrays/inst/doc/arrays.html) for using Bioconductor to analyze microarray data

```{r setup}
# set the working directory to `data` folder that is parallel to the `analysis` folder, which contains
# this script
knitr::opts_knit$set(root.dir = normalizePath("../data")) 
```

# Install required packages
| Package Name | Use |
|--------------|-----|
| affy | read and preprocess CEL files |
| limma | differential gene expression |
| annotate | load probe set annotation |
| ygs98.db | the microarray design file |


```{r load_libraries}
require(ggplot2)

# --- uncomment the code below for installing ---
# install affy and limma if it is not available 
# by uncommenting the following lines of code.
# recomment them after you have them installed
# 
# if (!requireNamespace("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#
# BiocManager::install(c("affy","limma"))
# BiocManager::install(c("annotate","ygs98.db"))
#
# install statmod for robust linear fitting
# install.packages("statmod")

require(affy)
require(limma)
require(annotate)
require(ygs98.db)
```

# Read data
```{r read_data}
# set the target file destination
URL <- "https://www.ebi.ac.uk/arrayexpress/files/E-GEOD-8536"
SDRF.file <- "E-GEOD-8536.sdrf.txt"
Data.file <- "E-GEOD-8536.raw.1.zip"

# download data
# ---
# uncomment the following lines to download the data
# once downloaded, recomment the lines
# ---
# download.file(paste(URL,SDRF.file,sep="/"), SDRF.file)
# download.file(paste(URL,Data.file,sep="/"), Data.file)
# unzip(Data.file)

# import metadata
SDRF <- read.delim("E-GEOD-8536.sdrf.txt",check.names=FALSE,stringsAsFactors=FALSE)
SDRF <- SDRF[order(SDRF$`Array Data File`),] # order the meta info table by scan names
SDRF[,c("Array Data File","Scan Name")]

# import and inspect array data
eset <- justRMA(filenames = SDRF$`Array Data.file`)
```

# Formatting data
```{r formatting_data}
# create a vector that will be used to assign experiment names to the eset object
scanName <- SDRF$`Scan Name`; names(scanName) <- SDRF$`Array Data File`

# now let's try to extract the time point
tmp <- do.call(rbind, strsplit(SDRF$`Scan Name`, " ")); rownames(tmp) <- SDRF$`Array Data File`
timePoint <- factor(tmp[colnames(eset),1], levels=c(1,12,24,48,60,120,340))

# convert the colnames of the eset object from array file name to scan name (=experimental condition)
# colnames(eset) <- scanName[colnames(eset)]
head(exprs(eset))
plotMDS(eset, cex = 0.8, labels = tmp[colnames(eset),1], col = colorRamp(c("yellow","blue"))(1:6/6)[timePoint])

# gene annotation
ID <- featureNames(eset) # extract the probe IDs
# you can decide what information to extract
# see `columns(ygs98.db)` for possible column names
# the following line is commented out because the "description" column is too long. however, it may be
# useful for digging deeper into the functions of the genes
# Symbol <- select(ygs98.db, keys = ID, columns = c("ORF", "GENENAME", "ENTREZID", "DESCRIPTION"))
Symbol <- select(ygs98.db, keys = ID, columns = c("ORF", "GENENAME"))
fData(eset) <- Symbol # this assigns the annotation info to the feature Data of the eset object

# remove probes that do not have an ORF associated
hasORF <- !is.na(fData(eset)$ORF)
eset1 <- eset[hasORF,]
dim(eset1)
```

# Perform differential gene expression

```{r DGE}
design <- model.matrix(~timePoint)
fit <- lmFit(eset1, design)
fit <- eBayes(fit, trend=TRUE, robust=TRUE)
res <- decideTests(fit)
summary(res)

# look at the top induced genes at 12 hour
topTable(fit, coef = "timePoint12", n = 10, adjust.method = "BH")

# plot the expression level changes of the top gene across timepoints
probeID <- topTable(fit, coef = "timePoint12", n = 10, adjust.method = "BH")[,c("PROBEID","ORF")]
expr <- exprs(eset1)[probeID$PROBEID,]; expr = c(t(expr)) # rotate the expression matrix and turn it into a vector
# now it is ordered such that each probe goes from timepoint 1-7, then the next gene
exdat <- data.frame(timePoint = rep(timePoint, times = 10), ORF = rep(probeID$ORF, each = 21), expr = expr)
p <- ggplot(exdat, aes(x = timePoint, y = expr)) + geom_point() + facet_wrap(~ ORF)
p
```
_Discussion_

1. Here each timepoint from 12 hr to 340 hr is compared to the first timepoint.
1. The "up" category in the "Intercept" column is not to be interpreted as "upregulated". This simply means the intercept is significantly above zero, that is, the gene is expressed at time zero. This is still a bit surprising though, as I would expect some genes not detectable at any condition.
1. Judging from the number of significantly up and down genes at each time point, I can't easily tell if there is any pattern, e.g. more gene expression changes at latter time points. But since this is the first pass of the analysis, I need to look more behind the curtain to see what's going on with individual genes.