---
title: reproduce Fig. 2 from Rasko et al 2005
date: 2020-05-08
author: Bin He
output: html_notebook
---

```{r setup}
# set the working directory of this notebook to the output folder
# that's where the input files are and where the figures will be stored
knitr::opts_knit$set(root.dir = normalizePath("../output"))
library(tidyverse)
```


First we will read in the data with our custom blast output format
For details, see `../script/run-blastp.sh`
```{r load_data}
# blast tabular output default column names
blast.outfmt7.col.names <- 
  c("qseqid","sseqid","qlen","slen","pident","mismatch","score","bitscore","evalue")
ref <- read_tsv("ref-self-blast.txt", comment = "#", col_names = blast.outfmt7.col.names)
query1 <- read_tsv("ref-query1-blast.txt", comment = "#", col_names = blast.outfmt7.col.names)
query2 <- read_tsv("ref-query2-blast.txt", comment = "#", col_names = blast.outfmt7.col.names)
```

Now we will calculate the Blast score ratios
```{r calc_bsr}
# 1. for each reference peptide, find the highest score in each of the three blast output and record them as Reference, Query1 and Query2
refscore <- ref %>% 
  group_by(qseqid) %>% 
  summarize(score.ref = max(score))

q1score <- query1 %>% 
  # the mutate statement removes extra characters in sseqid
  mutate(sseqid = str_sub(sseqid, 4, -2)) %>% 
  group_by(qseqid) %>% 
  select(qseqid, sseqid, score) %>%
  top_n(n = 1, wt = score)

q1score %>% 
  count(qseqid) %>% 
  arrange(desc(n))
```

Apparently some of the peptides in the reference genome have multiple hits in the query proteome with the same (highest) score.

I manually examined some of the instances and discovered that the corresponding subject sequences in the query proteome are identical despite different seq ids.

Nonetheless, we will carry on for the moment.

```{r calc_bsr_2}
q1score <- query1 %>% 
  group_by(qseqid) %>% 
  summarize(score.query1 = max(score))

q2score <- query2 %>% 
  group_by(qseqid) %>% 
  summarize(score.query2 = max(score))

bsr1 <- refscore %>% 
  left_join(q1score, by = "qseqid", suffix = c("ref","query1")) %>% 
  mutate(bsr1 = score.query1/score.ref)

bsr2 <- refscore %>% 
  left_join(q2score, by = "qseqid", suffix = c("ref","query2")) %>% 
  mutate(bsr2 = score.query2/score.ref)

bsr <- full_join(bsr1, bsr2, by = "qseqid") %>% select(qseqid, bsr1, bsr2)
```

Now we have all the data, and we can start plotting!
```{r plottng}
# to reproduce the scatter plot comparing the two query genomes
bsr %>% ggplot(aes(x = bsr1, y = bsr2)) + geom_point()

# to immitate the published figure, let's add some colors
th = 0.4 # this is how they define the four quadrants
# now let's assign each dot to one of the quadrants
bsr.col <- bsr %>% 
  mutate(class = factor((bsr1 > th)*1 + (bsr2 > th)*2, levels = 0:3, labels = c("unique to reference","specific to G9241", "specific to 10987", "conserved")))
# now let's try again, with some colors
colors <- c("orange3","green4","blue4","red3")
bsr.col %>% ggplot(aes(x = bsr2, y = bsr1, col = class)) + 
  geom_point(cex = 0.5) + scale_color_manual(values = colors) +
  xlab("B. cereus ATCC10987") + ylab("B. cereus G9241")
ggsave("figure2-reproduce.pdf")
```

